nextflu: Real-time tracking of seasonal influenza virus evolution in humans

### 2015-05-22 manuscript/correspondence/response.tex

\vspace{1cm}

Dear editor,

Please find attached our revised manuscript entitled ```nextflu: Real-time tracking of seasonal influenza virus evolution in humans''.  This is a resubmission of manuscript number BIOINF-2015-0560.

We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed poin-by-point response below. We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.

@@

\subsection*{Comments to the Author}

Neher and Bedford report ``nextflu'', an interactive near real time web interface to browse and explore the most informative approx. 2000 influenza HA gene sequences deposited in GISAID for the four most common strains. The automated pipeline ``augur'' fetches most recent data from GISAID, prunes the sequences which are subsequently combined to one multiple alignment with MAFFT. From this alignment, a phylogenetic tree is constructed and additional information is inferred, both of which are displayed on the interactive web site via ``auspice''.

The freely available Python source code is of exemplary structure and quality, the website is well documented and clearly structured. Nevertheless, I want to raise some critical comments regarding the application note as well as the application itself:

{\bf We would like to thank the reviewer for the supportive and constructive comments.}

@@

\subsection*{Manuscript}

1. Despite the fact that I do not fully understand the computation of the ``frequency trajectories of mutations'' I suspect that is tries to express the same measure as the ``frequencies of alleles'' in the ``Allele dynamic plots'' of Steinbrueck \& McHardy (NAR 2011) with the addition of relating it to a noise model. You should include a statement of comparison to AD  plots in your manuscript.

{\bf yes, the mutation trajectories are essentially the same, but nextflu also provides frequiencies of clades in the tree. The trajectories provide a useful sense of the temporal change. We have included a reference to the suggested paper. TODO}

@@

2. The prefix ``next'' of ``nextflu'' raises hope that the application will provide some forecast or recommendation which influenza might be the next to be battled with vaccination. This hope is further fueled by the expression ``predictors of successful clades''. However, I cannot see how to obtain such predictions with nextflu. Also the frequency plot does not seem to state something about future strain frequencies.

{\bf The purpose of nextflu is to integrate as much information as possible to get an overview of the currently circulating viruses. As such, nextflu can help making an informed vaccine strain selection. We included quantities that have been suggested to be predictive for this purpose. However, we feel that predictors are not mature or specific enough to make an automated prediction and many practical constraints (virus/serum availability, cell-culture behavior etc) needs to be taken into account when selecting the vaccine strain. By providing the predictors along with explicit temporal dynamics of clades in the tree, the user can evaluate their relative merits and pick the most promising strain, given other constraints.}

@@

3. The use case of the application is not clear to me. Can you give a more explicit example how information retrieval with nextflu could foster selection of the ``next successful clade'', rather than browsing the plain GISAID sequences? Of course, nextflu just visualizes existing data and thus cannot make up something that could not also be found in the database itself, but it would be nice to see how you intend to use nextflu.

{\bf Nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants. We have received enthusiastic feedback from the influenza community}

@@

4. Please give some hints why ``nextflu'' is not real-time but only near-real-time.

{\bf As of now, nextflu requires manual update which we do on a weekly basis. Future integration into GISAID might make this step obsolete. Furthermore, nextflu can only track sequences in the data base, and as they are deposited with some lag, nextflu isn't traking influenza in real time -- strictly speaking.}

@@

6. Typo: ``by laboratories around to world'' $\rightarrow$ ``by laboratories around the world''

\textbf{Fixed.}

7. You report that HA genes from GISAID are downloaded. Is the alignment build by whole genes or just the CDS?

{\bf We use the entire HA segment, but include shorter sequences if necessary.}

@@

8. Please explain what the variable $i$ stands for.

{\bf We have worked extensively on this section, see below.}

@@

{\bf We provided a method tab in nextflu.}

\subsection*{Application}

1. You give information about the workflow of the augur pipeline to process data of the underlying GISAID database. What happens if GISAID is updated. Do you have to manually restart augur? Is it possible to trigger such an update by the user of nextflu? If it is not possible, is the term ``real-time'' really satisfied?

{\bf as of now, we need to update manually. In any case, the deposition of sequences incurs a greater lag than our update frequency. We settled for ``near real time''.}

@@

2. You describe the data pruning of the GISAID sequences, which also involves a step of pruning the alignment to various criteria. Is this done automatically? It would be nice to give the user the capability to parameterize some of these criteria prior to visualizing the tree.

{\bf Since building aligning $>2000$ sequences and building a tree requires substantial computation, we can't have the user choose sequence filtering criteria and rerun the pipeline on demand. To address this concern, we have included multiple time resolutions. Users who want to customize this further can always clone the github repo and run the pipeline locally.}

@@

3. It is rather complicated to zoom in and out or to scroll in the tree. Would it be possible to add according navigation buttons, rather then resetting the whole display when zooming out a bit?

{\bf With the current set-up, this is not readily achieved.}

@@

4. It is difficult to use the date slider to actually move through time since the refresh is only accomplished when releasing the mouse button. Also slight horizontal movements can cause unwanted jumps. Could you maybe provide two buttons to jump to the neighboring date points? Or a functionality to automatically move through time like in a movie.

{\bf Dynamic update of the tree imposes a heavy computational burden. To address this issue, we now update a limited set of features as the slider is moved. We feel that the current compromise of updating tip circles without recalculating predictors and colors is a good compromise. Depending on the speed of the user's computer, this well be more or less smooth.}

@@

\subsection*{Minor revisions}

p1, col1, l47: ``near real-time'', can the authors specify the updating frequency? from the website, it seems to be every 5 days.

p1, col2, l54: how is the outgroup chosen?

{\bf We choose a well characterized strain predating the time interval of interest by a few years. We updated the text to clarify this.}

p2, col1, l21: typo: the sampling time point is $t_v$ not $x(t_v)$
{\bf Fixed, thank you.}

@@

%%% REVIEWER 3 %%%
\section*{Reviewer 3}

\subsection*{Major comments}

This may be beyond the scope of the tool, but it would be extremely useful if the tool could visualize not just the \# of epitope or receptor binding site changes, but which specific amino acid changes are involved -- at least for the branches defining the main clusters identified (e.g., 3c3.a and 3c2.a).

{\bf Tool tips of branches now show the amino-acid mutations associated with the branch. Furthermore, the we added a display of all alignment variability that allows to spot variable positions. Clicking on these positions in the graph will plot the frequencies of amino acids and color the tree accordingly. }

@@

\subsection*{Minor comment}

Could the authors clarify what they mean by `removing known reassortant clusters'?  This is very vague and I assume naturally occurring reassortant viruses are not being removed.

{\bf We filter sequences that are very similar to sequences from outbreaks of triple reassortants with pH1N1. ADD DETAIL.}

### 2015-05-22 manuscript/methods.tex
continued working on response

\documentclass{bioinfo}
\copyrightyear{2015}
\pubyear{2015}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amsmath}

\newcommand{\EQ}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\EQS}[2]{Eqs.~(\ref{eq:#1}) and (\ref{eq:#2})}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\TAB}[1]{Tab.~\ref{tab:#1}}
\newcommand{\REF}[1]{ref.~\citep{#1}}
\newcommand{\augur}{\textbf{augur}}
\newcommand{\auspice}{\textbf{auspice}}
\newcommand{\nextflu}{\textbf{nextflu}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Tracking of seasonal influenza virus evolution]{nextflu: Real-time tracking of seasonal influenza virus evolution in humans}
\author{Richard~A.~Neher$^{1}$ and Trevor Bedford$^{2}$}
\address{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany, and $^{2}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

%\date{\today}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly, allowing them to evade immunity in their human hosts and reinfect previously infected individuals.
Similarly, vaccines against seasonal influenza need to be updated frequently to protect against an evolving virus population.
We have thus developed a processing pipeline and browser-based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
This web-application displays a phylogenetic tree that can be decorated with additional information such as the viral genotype at specific sites, sampling location, and derived statistics that have been shown to be predictive of future virus dynamics.
Additionally, mutation, genotype and clade frequency trajectories are calculated and displayed.

\section{Availability and implementation:} Python and Javascript source code is freely available from \url{https://github.com/blab/nextflu}, while the web-application is live at \url{http://nextflu.org}.

\section{Contact:} tbedford@fredhutch.org

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{Introduction}

Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
Vaccination remains the most effective public health measure to combat seasonal epidemics.
However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.

\begin{figure}[t!]
	\centering
	\includegraphics[width=0.99\columnwidth]{tree}
	\caption[]{The \nextflu{} website with the user interface on the left and the phylogenetic tree on the right.}
	\label{fig:tree}
\end{figure}

In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
The first step in the processing pipeline is to automatically select a subset of representative viruses.
Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.

From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
Next, the state of every internal node of the tree is inferred using a marginal maximum likelihood method and missing sequence data at phylogeny tips is filled with the nearest ancestral sequence at these sites.
Internal branches without mutations are collapsed into polytomies.
The final tree is decorated with the attributes to be displayed in the browser. %, as described below.

In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
The frequencies $x_i$ corresponding to a feature $\phi$ (e.g.\ mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
\begin{equation}
	\label{eq:obs}
	f(\mathbf{x}) = \prod_{v\in V_\phi} x(t_v) \prod_{v\notin V_{\phi}} (1-x(t_v))
\end{equation}
where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $x(t_v)$ is the sampling timepoint of virus $v$, and the process noise model
\begin{equation}
	\label{eq:freq}
	g(\mathbf{x}) = \mathrm{exp}\left(
		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \Delta t_i}
	\right)
\end{equation}
where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the  change in interval $i-1$.
$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
We typically use $\epsilon = 0.7$.
Here, the objective function $\mathrm{log}(f) + \mathrm{log}(g)$ is maximized using SciPy's implementation of a downhill simplex algorithm \citep{oliphant_python_2007}.
To speed up convergence, optimization is first done using a coarse grid of pivot points which is subsequently refined to 12 pivots per year. Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.
At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.

These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
\begin{itemize}
	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
%	\item non-epitope mutations: 
%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
 %   \item geographic location
    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
\end{itemize}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.99\columnwidth]{frequencies}
	\caption[]{The frequency diagram allows geography-specific plotting of frequencies of individual mutations, pairs of mutations and clades in the tree.}
	\label{fig:freq}
\end{figure}

The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.

%\section{Conclusion}
We built \nextflu{} to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around the world.
By using the most recent data and integrating phylogenies with frequency trajectories and predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines. 
\nextflu{} was designed to be readily adapted to other rapidly evolving viruses and we see significant room for future developments in this area.

\paragraph{Funding\textcolon}This work is supported by the ERC though Stg-260686 and by the NIH through U54 GM111274.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{}

\bibitem[Bogner {\em et~al.}, 2006]{GISAID}
Bogner,P. {\em et~al.} (2006{\em{}}) A global
  initiative on sharing avian flu data.
\newblock {\em Nature, } {\bf 442} (7106), 981--981.

\bibitem[Bostock {\em et~al.}, 2011]{bostock_d3_2011}
Bostock,M. {\em et~al.} (2011{\em{}}) D$^3$: data-driven
  documents.
\newblock {\em IEEE Trans Vis Comput Graphics, } {\bf 17} (12), 2301--2309.

\bibitem[Katoh and Standley, 2013]{katoh_mafft_2013}
Katoh,K. and Standley,D.M. (2013{\em{}}) {MAFFT} multiple sequence alignment
  software version 7: improvements in performance and usability.
\newblock {\em Mol Biol Evol, } {\bf 30}, 772--780.

\bibitem[Koel {\em et~al.}, 2013]{koel_substitutions_2013}
Koel,B.F. {\em et~al.} (2013{\em{}})
  Substitutions near the receptor binding site determine major antigenic change
  during influenza virus evolution.
\newblock {\em Science, } {\bf 342} (6161), 976--979.

\bibitem[\L{}uksza and L\"a{}ssig, 2014]{luksza_predictive_2014}
\L{}uksza,M. and L\"a{}ssig,M. (2014{\em{}}) A predictive fitness model for
  influenza.
\newblock {\em Nature, } {\bf 507} (7490), 57--61.

\bibitem[Neher {\em et~al.}, 2014]{neher_predicting_2014}
Neher,R.A. {\em et~al.} (2014{\em{}}) Predicting evolution
  from the shape of genealogical trees.
\newblock {\em eLife Sciences, } {\bf 3}, e03568.

\bibitem[Oliphant, 2007]{oliphant_python_2007}
Oliphant,T. (2007{\em{}}) Python for scientific computing.
\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.

\bibitem[Price {\em et~al.}, 2009]{price_fasttree_2009}
Price,M.N. {\em et~al.} (2009{\em{}}) {FastTree:} computing
  large minimum evolution trees with profiles instead of a distance matrix.
\newblock {\em Mol Biol Evol, } {\bf 26} (7), 1641--50.

\bibitem[Stamatakis, 2014]{stamatakis_raxml_2014}
Stamatakis,A. (2014{\em{}}) {RAxML} version 8: a tool for phylogenetic analysis
  and post-analysis of large phylogenies.
\newblock {\em Bioinformatics, } {\bf 30} (9), 1312--1313.

\bibitem[{World Health Organization}, 2009]{flufactsheet}
{World Health Organization} (2009{\em{}}) {\em {Influenza Fact sheet}}.
\newblock Available at http://www.who.int/mediacentre/factsheets/fs211/en/.

\end{thebibliography}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\title{nextflu.org methods summary}

\section{Pruning and alignment}
The first step in the processing pipeline is to automatically select a subset of representative viruses.
Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters. As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.

\section{Tree building}
From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
Next, the state of every internal node of the tree is inferred using a marginal maximum likelihood method and missing sequence data at phylogeny tips is filled with the nearest ancestral sequence at these sites.
Internal branches without mutations are collapsed into polytomies.
The final tree is decorated with the attributes to be displayed in the browser.

\section{Frequency calculation}
\augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
The frequencies $x_i$ corresponding to a feature $\phi$ (e.g.\ mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
\begin{equation}
	\label{eq:obs}
	f(\mathbf{x}) = \prod_{v\in V_\phi} x(t_v) \prod_{v\notin V_{\phi}} (1-x(t_v))
\end{equation}
where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $t_v$ is the collection date of virus $v$, and the process noise model
\begin{equation}
	\label{eq:freq}
	g(\mathbf{x}) = \mathrm{exp}\left(
		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \Delta t_i}
	\right)
\end{equation}
where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the  change in interval $i-1$.
$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
We typically use $\epsilon = 0.7$.
Here, the objective function $\mathrm{log}(f) + \mathrm{log}(g)$ is maximized using SciPy's implementation of a downhill simplex algorithm \citep{oliphant_python_2007}.
To speed up convergence, optimization is first done using a coarse grid of pivot points which is subsequently refined to 12 pivots per year. Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.

\section{References}

### 2015-05-22 manuscript/nextflu.tex
continued working on response

Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
Vaccination remains the most effective public health measure to combat seasonal epidemics.
However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.

@@

In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
The first step in the processing pipeline is to automatically select a subset of representative viruses.
Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.

From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
Next, the state of every internal node of the tree is inferred using a marginal maximum likelihood method and missing sequence data at phylogeny tips is filled with the nearest ancestral sequence at these sites.
Internal branches without mutations are collapsed into polytomies.
The final tree is decorated with the attributes to be displayed in the browser. %, as described below.

In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
The frequencies $x_i$ corresponding to a feature $\phi$ (e.g.\ mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
\begin{equation}
	\label{eq:obs}
	f(\mathbf{x}) = \prod_{v\in V_\phi} x(t_v) \prod_{v\notin V_{\phi}} (1-x(t_v))
\end{equation}
where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $x(t_v)$ is the sampling timepoint of virus $v$, and the process noise model
\begin{equation}
	\label{eq:freq}
	g(\mathbf{x}) = \mathrm{exp}\left(
		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \Delta t_i}
	\right)
\end{equation}
where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the  change in interval $i-1$.
$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
We typically use $\epsilon = 0.7$.
Here, the objective function $\mathrm{log}(f) + \mathrm{log}(g)$ is maximized using SciPy's implementation of a downhill simplex algorithm \citep{oliphant_python_2007}.
To speed up convergence, optimization is first done using a coarse grid of pivot points which is subsequently refined to 12 pivots per year. Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.
At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.

These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
\begin{itemize}
	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
%	\item non-epitope mutations: 
%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
 %   \item geographic location
    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
\end{itemize}

@@

The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.

>>>>

Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
Vaccination remains the most effective public health measure to combat seasonal epidemics.
However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data. By visualizing many different observables and predictors, we hope that \nextflu{} will help vaccine strain selection.
Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.

@@

In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
The first step in the processing pipeline is to automatically select a subset of representative viruses.
Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters. As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.

From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
Next, the state of every internal node of the tree is inferred using a marginal maximum likelihood method and missing sequence data at phylogeny tips is filled with the nearest ancestral sequence at these sites.
Internal branches without mutations are collapsed into polytomies.
The final tree is decorated with the attributes to be displayed in the browser.

In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. Frequency are determined by maximizing the likelihood of sampling the observed set of virus sequences. In addition, we impose a smoothing that penalized rapid changes in frequency of the frequency derivative. \augur{} estimates frequency with up to one month resolution using a downhill simplex algorithm \citep{oliphant_python_2007}. At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.

These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
\begin{itemize}
	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
%	\item non-epitope mutations: 
%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
 %   \item geographic location
    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
\end{itemize}
The tree display can also be restricted to different geographic regions.

@@

The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted. A second plot shows the variability of the alignment. On mouse-click on a variable positions in this plot, \auspice{} will color the tree by amino-acid at this position and plot the frequencies mutations at his positions.
